Co-methylation

### 2014-10-03 Rmarkdown/comethylation.Rmd
Reviewed within-fragment co-methylation of Cokus (2008).

@@

* More recently, within-fragment analyses. Describe concept (move from next section to this section).
* Cokus
* \cite{Lacey:2009by}
* \cite{Paliwal:2013fk}

@@

@@

Within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

As every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads, that is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

@@

If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci. However, current technologies for bisulfite-sequencing mean that we are limited to studying reads that, at most, span about $200-250$ bp and so are limited to studying co-methylation of methylation loci that are very close to one another along the chromosome. Obviously, we cannot study within-fragment co-methylation from microarray data since these produce measurements of methylation that are aggregated across many hundreds or thousands of cells.

>>>>

#### \cite{Cokus:2008fc}\label{sec:Cokus}

@@

Using bisulfite-sequencing technologies we can analyse co-methylation at the level of individual DNA molecules. I call this within-fragment co-methylation, where fragment refers to a fragment of DNA. More precisely, within-fragment co-methylation is the dependence of DNA methylation at methylation loci that occur on the same DNA fragment. It can be thought of as the dependence structure of the binary stochastic process $Z_{h} = (Z_{h, 1}, Z_{h, 2}, \ldots, Z_{h, N_{loci}})$. For example, we may wish to understand how $Z_{h, i}$ affects $Z_{h, i'}$ and to develop a simple, tractable model of this dependence structure. We can study what variables effect this dependence, such as the distance between methylation loci and the genomic context of the methylation loci.

Because every sequencing read is produced from a single DNA fragment[^chimera], we can study within-fragment co-methylation by analysing the methylation patterns along individual sequencing reads. That is, by studying the methylation patterns at m-tuples. Furthermore, since the DNA methyltransferases act on individual DNA fragments (__TODO: check and cite__), these estimates of within-fragment co-methylation are on the same scale as the physical process that lays down and maintains DNA methylation.

If we could sequence entire chromosomes by single reads then we could analyse co-methylation between any set of methylation loci. However, the most commonly used technology for high-throughput bisulfite-sequencing, which uses Illumina's  sequencing technology, can only generate reads that span approximately $200-250$ bp at a maximum. Assembling these reads into longer contigs will not solve this problem since this cannot guarantee that the assembled reads actually originated from the same DNA fragment.

Long-read sequencing technologies, such as Pacific Biosciences (__CITE__) and Oxford Nanopore (__CITE__), would overcome this limitation but the lesser throughput of these technologies and the reliability with which they can identify methylcytosines create new complications (__CITE__). An alternative solution, although completely hypothetical, is to use a technique that generates synthetic long reads, similar to those generated by Illumina's TruSeq Synthetic Long-Read technique[^moleculo](__CITE__).

[^moleculo]: My understanding is that Illumina's TruSeq Synthetic Long-Read technique cannot be directly applied to bisulfite-sequencing for a number of technique reasons. These include the fragility of large DNA fragments treated with sodium bisulfite (__TODO: Discuss with someone from AGRF__)

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200$ nucleotides. There is less previous work studying within-fragment co-methylation than there is for correlations of aggregate methylation levels. Here I review this previously published work.

As already discussed in Section \ref{sec:Cokus}, \cite{Cokus:2008fc} identify a $10$ nucleotide periodicity in the autocorrelation of $\beta$-values for CHH methylation from a whole-genome bisulfite-sequencing experiment of _Arabidopsis Thaliana_. They find this same periodicity "when individual reads [were] examined directly".

The sequencing technology at the time of this publication produced very short reads, on average only $31$ nucleotides of usable sequence. For the results shown in Supplementary Figure 14 of \cite{Cokus:2008fc}, they first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \cite{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. This plot includes pointwise $95\%$ confidence intervals, computed under an assumption of Normality, and a running average of these averages (). There is an approximate $10$ nucleotide periodicity evident in this plot, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

[^cokus_clarification]: Actually, it's not clear whether they consider all pairs of CHH loci or only the first pair of CHH loci in a read since a read may contain multiple pairs. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

I consider these results to use a restricted definition of within-fragment "co-methylation". These results tells us nothing about the probability that both CHH loci are _unmethylated_ nor the joint probability distribution of methylation for two CHH loci on the same fragment. Specifically, \cite{Cokus:2008fc} only estimate $Pr(Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$, where $Z_{h, i}$ and $Z_{h, i'}$ are two CHH loci on the same fragment $(i < i')$. From these two quantities we can only estimate two of the four joint probabilities, $Pr(Z_{h, i} = x, Z_{h, i} = y)$, namely, $Pr(Z_{h, i} = 1, Z_{h, i'} = 1)$ and $Pr(Z_{h, i} = 1, Z_{h, i'} = 0)$. It does not allow estimation of $Pr(Z_{h, i} = 0, Z_{h, i'} = 0)$, nor of $Pr(Z_{h, i} = 0, Z_{h, i'} = 1)$, since these require estimates of $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$. While it is trivial to additionally estimate $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$, I argue in Section \ref{subsec:wf_cometh} that it is more useful and simpler to  estimate the odds ratio, $\frac{Pr(Z_{h, i} = 1, Z_{h, i'} = 1) Pr(Z_{h, i} = 0, Z_{h, i'} = 0)}{Pr(Z_{h, i} = 1, Z_{h, i'} = 0) Pr(Z_{h, i} = 0, Z_{h, i'} = 1)}$, rather than the conditional probabilities, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$ and $Pr(Z_{h, i'} = 0 | Z_{h, i} = 0)$.

Furthermore, by aggregating all pairs of CHH loci separated by a common distance and then estimating these probabilities, there is an implicit assumption that all such pairs on the same chromosome have the same conditional probability, $Pr(Z_{h, i'} = 1 | Z_{h, i} = 1)$. There is good reason to suspect that this is not true and that there exists significant pair-to-pair variability. For example, we know for CpGs that those within CpG islands have a vastly different probability of being methylated than do those outside of islands; it is reasonable to suspect that this is also true for CHH methylation in different regions of the _Arabidopsis_ genome and hence for "co-methylation" of these same loci.

__PREVIOUS WORK__

* \cite{Lacey:2009by}
* \cite{Paliwal:2013fk}

@@

__TODO: A spiel on how my method is better than existing ones.__

@@

__TODO: A spiel on how my method is better than existing ones.__

@@
